% Copyright 2018 Google LLC
%
% Use of this source code is governed by an MIT-style
% license that can be found in the LICENSE file or at
% https://opensource.org/licenses/MIT.

%!TeX spellcheck = en-US

\documentclass[eprint.tex]{subfiles}
\begin{document}
\section{Introduction}
Two aspects of disk encryption make it a challenge for cryptography.  First,
performance is critical; every extra cycle is a worse user experience, and on a mobile device
a reduced battery life.  Second, the ciphertext can be no larger than the plaintext: a sector-sized
read or write to the filesystem must mean a sector-sized read or write to the underlying device,
or performance will again suffer greatly
(as well as, in the case of writes to flash memory, the life of the device).
Nonce reuse is inevitable as there is nowhere to store a varying nonce, and there is no space
for a MAC; thus standard constructions like AES-GCM are not an option and standard notions
of semantic security are unachievable.  The best that can be done under the circumstances is
a ``tweakable super-pseudorandom permutation'': an adversary with access to both encryption
and decryption functions who can choose tweak and plaintext/ciphertext freely is unable to
distinguish it from a family of independent random permutations.

\subsection{History}

Hasty Pudding Cipher~\cite{hpc} was a variable-input-length primitive presented to the AES contest.
A key innovation
was the idea of a ``spice'', which was later formalized as a ``tweak'' in~\cite{tweakable}.
Another tweakable large-block primitive was Mercy~\cite{mercy},
cryptanalyzed in~\cite{mercycryptanalysis}.

\cite{luby-rackoff} (see also~\cite{maurer-luby-rackoff,ppdes})
shows how to construct a pseudorandom permutation using a three-round Feistel
network of pseudorandom functions;
proves that this is not a secure super-pseudorandom permutation (where the adversary
has access to decryption as well as encryption) and that four rounds suffice for this aim.
BEAR and LION~\cite{bearlion} apply this result to an unbalanced Feistel network to build a
large-block cipher from a hash function and a stream cipher (see also BEAST~\cite{beast}).

\cite{fasterlr} shows that a universal function
suffices for the first round, which~\cite{NaorReingold} extends to four-round function
to build a super-pseudorandom permutation.

More recently, proposals in this space have focused on the use of
block ciphers. VIL mode~\cite{brvil} is a CBC-MAC based two-pass variable-input-length construction which
is a PRP but not an SPRP.
SPC mode~\cite{bd99} extends this to an SPRP.
\cite{cmc} gives the concrete security definition for
a tweakable, variable-length, length-preserving SPRP,
and defines CMC mode, a two-pass mode with a quadratic security bound.
EME mode~\cite{eme} is similar but parallelizable, while
EME* mode~\cite{emestar} extends EME mode to handle blocks that are not a multiple of the block
cipher size. PEP~\cite{pep}, TET~\cite{tet}, and HEH~\cite{heh} have a mixing layer either side of
an ECB layer.

XCB~\cite{xcb} is a block-cipher based unbalanced three-round Feistel network with an
$\epsilon$-almost-XOR-universal hash function for the first and third rounds
(``hash-XOR-hash''),
which uses block
cipher invocations on the narrow side of the network to ensure that the network is an SPRP, rather
than just a PRP.
HCTR~\cite{hctr,hctr2,kumarhctr}, HCH~\cite{hch}, HSE~\cite{hse}, and HMC~\cite{hmc} reduce this to a single
block cipher invocation within the Feistel network.
These proposals require
either two AES invocations, or an AES invocation and two $\GF(2^{128})$ multiplications,
per 128 bits of input.

\subsection{Our contribution}
On the ARM architecture, the ARMv8 Cryptography Extensions include instructions that make
AES and $\GF(2^{128})$ multiplications much more efficient. However,
smartphones designed for developing markets
often use lower-end processors which
don't support these extensions, and as a result there is no existing SPRP construction which performs
acceptably on them.

On such platforms stream ciphers such as ChaCha12~\cite{chacha} significantly
outperform block ciphers in cycles per byte, especially with constant-time implementations.
Similarly, absent specific processor support, hash functions such as NH~\cite{umac2} and
Poly1305 hash~\cite{poly1305} will be much faster
than a $\GF(2^{128})$ polynomial hash. Since these are the operations that act on the bulk of
the data in a disk-sector-sized block, a hash-XOR-hash
mode of operation relying on them should achieve
much improved performance on such platforms.

To this end, we present the HBSH (hash, block cipher, stream cipher, hash)
construction, which generalizes over constructions such as
HCTR and HCH by taking an $\epsilon$-almost-$\Delta$-universal hash function and a
nonce-accepting stream cipher
as components. Based on this construction, our main proposal is Adiantum,
which uses a combination of NH and Poly1305 for the hashing, XChaCha12 for the stream cipher, and
AES for the single block cipher application. Adiantum:
\begin{itemize}
    \item is a tweakable, variable-input-length, super-pseudorandom permutation
    \item has a security bound quadratic in the number of queries and linear in message length
    \item is highly parallelizable
    \item needs only three passes over the bulk of the data, or
        two if the XOR is combined with the second hash.
\end{itemize}

Without special cases or extra setup, Adiantum handles:
\begin{itemize}
    \item any message and tweak lengths within the allowed range,
    \item varying message and tweak lengths for the same keys.
\end{itemize}

We also describe HPolyC, which does not use NH.
HPolyC is 30\% slower on large blocks, but simpler and much more key agile.

\subbib
\end{document}
